Allocation Removal by Partial Evaluation in a Tracing JIT

### 2011-01-19 talk/pepm2011/presentation/talk.tex
a few visual bugs.
prize question: why does inkscape insert more and more empty lines into its svg
files?
feedback from the second training presentation

\usepackage[english]{babel}
\usepackage{listings}
\usepackage{ulem}
\usepackage{alltt}

\usepackage[utf8x]{inputenc}
% or whatever

@@

\begin{frame}
  \frametitle{What to do?}
  \begin{itemize}
      \item Use a JIT compiler
      \item \textbf{Add a optimization that can deal with heap operations}
      \pause
      \item optimize short-lived objects
      \item remove some of the redundancy
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Tracing JITs Compile by Observing an Interpreter}
  \begin{itemize}
      \item VM contains both an interpreter and the tracing JIT compiler
      \item JIT works by observing and logging what the interpreter does
      \item for interesting, commonly executed code paths
      \item produces a linear list of operations (trace)
      \item trace is optimized and then turned into machine code
  \end{itemize}
\end{frame}

@@

\begin{frame}[containsverbatim]
  \frametitle{Optimized Example Trace}
  Trace of \texttt{x = a + b; y = x + c}:
\begin{alltt}
guard_class(a, Integer)
guard_class(b, Integer)
i1 = get(a, intval)
i2 = get(b, intval)
i3 = int_add(i1, i2)
\sout{x = new(Integer)}
\sout{set(x, intval, i3)}
\end{alltt}
\begin{alltt}
\sout{guard_class(x, Integer)}
guard_class(c, Integer)
\sout{i4 = get(x, intval)}
i5 = get(c, intval)
i6 = int_add(\emph{i3}, i5)
y = new(Integer)
set(y, intval, i6)
\end{alltt}
\end{frame}

\begin{frame}
  \frametitle{The Advantages of Tracing JITs}
  \begin{itemize}
      \item Traces are interesting linear pieces of code
      \item most of the time correspond to loops
      \item everything called in the trace is inlined
      \item can perform good optimizations on the loop
      \item rarer paths run by the interpreter
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Optimizing the Heap Operations in a Trace}
  \begin{itemize}
      \item Contribution of our paper
      \item A simple, efficient and effective optimization of heap operations in a trace
      \item using online partial evaluation
      \item fully implemented and in use in large-scale interpreters
  \end{itemize}
  \pause
  \begin{block}{Ingredients}
      \begin{itemize}
          \item a slightly simplified model for objects on the heap
          \item operational semantics of trace operations that manipulate the heap
          \item optimization rules for those operations
      \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{Lifting}
  Problem: What happens if we write a static object into a dynamic one?
  \pause
  \begin{itemize}
      \item lose track of the static object because of possibility of aliasing
      \item need to \emph{lift} the static object
      \item lifting produces operations that recreate the static object
  \end{itemize}
\end{frame}

@@

\begin{frame}[containsverbatim]
  \frametitle{Comparison to "Dynamic Typing"/Boxing Analysis}
  \begin{itemize}
      \item those optimizations work ahead of time
      \item don't work for many dynamic languages, where the source simply does not contain enough information
  \end{itemize}

  \begin{block}{Python Example:}
  \begin{verbatim}
def sum(container, initial):
    result = initial
    for element in container:
        result = result + element
    return result
  \end{verbatim}
  \end{block}
\end{frame}

>>>>

\usepackage[english]{babel}
\usepackage{listings}
\usepackage{ulem}
\usepackage{color}
\usepackage{alltt}

\usepackage[utf8x]{inputenc}

\newcommand\redsout[1]{{\color{red}\sout{\hbox{\color{black}{#1}}}}}

% or whatever

@@

\begin{frame}
  \frametitle{What to do?}
  \begin{itemize}
      \item Use a JIT compiler
      \item \textbf{Add an optimization that can deal with heap operations}
      \pause
      \item optimize short-lived objects
      \item remove some of the redundancy
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Tracing JITs Compile by Observing an Interpreter}
  \begin{itemize}
      \item VM contains both an interpreter and the tracing JIT compiler
      \item JIT works by observing and logging what the interpreter does
      \begin{itemize}
          \item for interesting, commonly executed code paths
          \item produces a linear list of operations (trace)
      \end{itemize}
      \item trace is optimized and then turned into machine code
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The Advantages of Tracing JITs}
  \begin{itemize}
      \item Traces are interesting linear pieces of code
      \item most of the time correspond to loops
      \item everything called in the trace is inlined
      \item can perform good optimizations on the trace
      \item rarer paths run by the interpreter
  \end{itemize}
\end{frame}

@@

\begin{frame}[containsverbatim]
  \frametitle{Optimized Example Trace}
  Trace of \texttt{x = a + b; y = x + c}:
\begin{alltt}
guard_class(a, Integer)
guard_class(b, Integer)
i1 = get(a, intval)
i2 = get(b, intval)
i3 = int_add(i1, i2)
\redsout{x = new(Integer)}
\redsout{set(x, intval, i3)}
\end{alltt}
\begin{alltt}
\redsout{guard_class(x, Integer)}
guard_class(c, Integer)
\redsout{i4 = get(x, intval)}
i5 = get(c, intval)
i6 = int_add(\emph{i3}, i5)
y = new(Integer)
set(y, intval, i6)
\end{alltt}
\end{frame}

@@

\begin{frame}
  \frametitle{Contribution of our Paper}
  \begin{itemize}
      \item a simple, efficient and effective optimization of heap operations in a trace
      \item using online partial evaluation
      \item fully implemented and in use in large-scale interpreters
  \end{itemize}
  \pause
  \begin{block}{Ingredients}
      \begin{itemize}
          \item a slightly simplified model for objects on the heap
          \item operational semantics of trace operations that manipulate the heap
          \item optimization rules for those operations
      \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{Lifting}
  Problem: What happens if we write a static object into a dynamic one?
  \pause
  \begin{itemize}
      \item lose track of the static object because of possibility of aliasing
      \item need to \emph{lift} the static object
      \item lifting produces operations that recreate the static object
      \pause
      \item needs to be careful due to recursive structures
  \end{itemize}
\end{frame}

@@

\begin{frame}[containsverbatim]
  \frametitle{Comparison to "Dynamic Typing"/Boxing Analysis}
  \begin{itemize}
      \item those optimizations work ahead of time
      \item don't work for many dynamic languages, where the source simply does not contain enough information
  \end{itemize}
  \begin{block}{Python Example:}
  \begin{verbatim}
def sum(container, initial):
    result = initial
    for element in container:
        result = result + element
    return result
  \end{verbatim}
  \end{block}
\end{frame}
